Technique for \earlywarning\ filtering for compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-28 introduction.tex
addressed leos comments
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1173 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The gravitational wave signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup.  We have the ambition of reporting GW candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss a
real-time filtering algorithm for compact binary coalescence waveforms, which
is one aspect of the work required to make this possible.  We also present the
prospects for advanced \GW\ detectors and discuss other areas of work that
would be required for rapid analysis.

@@

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The time before coalescence that a source could be
detectable is a lower bound that does not include any latency inducing aspects
of a real analysis.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$. $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

@@

Here we provide a brief list of technical latencies associated with \CBC\
analysis in S6, VSR3 in chronological order \citep{HugheyGWPAW2011}.
%
\begin{itemize}
%
\item {\it Data calibration, quality control and distribution ($\sim
2$\,minutes):}  This is the time it took to calibrate the detector data, assess
its quality and distribute it to a central location for analysis.  Improvements
in other areas could make this the dominating source of latency. If possible
work should be done to improve this.
%
\item {\it Trigger production ($2-5$\,minutes):} This was an impressive latency
and was certainly sub-dominant in the S6, VSR3 analysis.  However, even if no
other sources of latency exist it is too long to catch prompt or extended
emission.  It is also more challenging for searches with advanced detectors
because the templates needed to filter the data are about 10 times longer.  We
will focus on reducing the latency of the trigger production step in this work.
%
\item {\it Event gathering, downselection and pointing calculations ($2-3$\,
minutes):} This was the time required to gather the event parameters and sky
localization from the various online analyses, downselect the events and
calculate pointings.  If other sources of latency were improved this could
dominate.  If possible work should be done to improve this.
%
\item {\it Human vetting of the trigger quality ($10-20$\,minutes):}  This was
by far the largest latency during the S6, VSR3 low latency analysis and will
hopefully be automated by the advanced detector era.  We therefore give it no
more consideration here.
%
\end{itemize}
%

In addition to the latencies described above, which may be overcome, there are
fundamental latencies associated with data acquisition and transfer that cannot
be overcome.  For example, the \LIGO{} data acquisition records science data in
1/16~s blocks~\citep{Bork2001}. Data must also be aggregated from all of the
\GW\ observatories using global computer networks capable of high bandwidth but
perhaps only modest latency.  This could introduce an additional latency of
$\sim$100~ms for a total of $\sim 1/4$\,s of unavoidable latency.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a \textsc{ns} is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The gravitational wave signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup.  We have the ambition of reporting GW candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss
one essential ingredient in this effort, a computationally inexpensive
real-time filtering algorithm for detecting inspiral signals in gravitational
wave data.  We also present the prospects for advanced \GW\ detectors and
discuss other areas of work that would be required for rapid analysis.

@@

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$. $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

@@

Here we provide a brief list of technical latencies associated with \CBC\
analysis in S6, VSR3 in chronological order \citep{HugheyGWPAW2011}.
%
\begin{itemize}
%
\item {\it raw data aggregation and broadcasting ($sim 0.25$\,s):}  There are
fundamental latencies associated with data acquisition and broadcasting that
cannot be overcome.  For example, the \LIGO{} data acquisition records science
data in 1/16~s blocks~\citep{Bork2001}. Data must also be aggregated from all
of the \GW\ observatories using global computer networks capable of high
bandwidth but perhaps only modest latency.  This could introduce an additional
latency of $\sim$100~ms.  These latencies are microscopic compared with the
other technical sources of latency below. 
%
\item {\it Data calibration, quality control and distribution ($\sim
2$\,minutes):}  This is the time it took to calibrate the detector data, assess
its quality and distribute it to a central location for analysis.  Improvements
in other areas could make this the dominating source of latency. If possible
work should be done to improve this.
%
\item {\it Trigger production ($2-5$\,minutes):} This was an impressive latency
and was certainly sub-dominant in the S6, VSR3 analysis.  However, even if no
other sources of latency exist it is too long to catch prompt or extended
emission.  It is also more challenging for searches with advanced detectors
because the templates needed to filter the data are about 10 times longer.  We
will focus on reducing the latency of the trigger production step in this work.
%
\item {\it Event gathering, downselection and pointing calculations ($2-3$\,
minutes):} This was the time required to gather the event parameters and sky
localization from the various online analyses, downselect the events and
calculate pointings.  If other sources of latency were improved this could
dominate.  If possible work should be done to improve this.
%
\item {\it Human vetting of the trigger quality ($10-20$\,minutes):}  This was
by far the largest latency during the S6, VSR3 low latency analysis and will
hopefully be automated by the advanced detector era.  We therefore give it no
more consideration here.
%
\end{itemize}
%
